Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-16 treestats_paper.tex
Spell check with EN_us.

\paragraph{Polarisation}
The node statistic defined above is appropriate for statistics such as nucleotide diversity, which is a function
of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
at a site, and we may define a \emph{polarised} node statistic without the second term from our previous definition:

\begin{align}
    \text{(polarised)} \qquad
    \nodep(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)) .
\end{align}

@@

In the definition above we sum over all alleles at each site.
However, sometimes it is useful to distinguish the \emph{ancestral} allele
(i.e., the allele at the root of the tree) from the remaining derived alleles.
This allows statistics in principle to differentiate ancestral from derived alleles,
information which is available in practice (albeit noisily),
and one way to make use of this information is to sum over only derived alleles.
Analogously to the above,
we say a site statistic is \textbf{polarised} if we do this,
defining
\begin{align} \label{eqn:site_polarised}
    \text{(polarised)} \qquad
    \sitep(f, \iw)_j
    &=
    \sum_{a \in D_j} f(\aw_j(a)) ,
\end{align}
where $D_j$ denotes the set of all alleles at site $j$ except the allele at the root of the tree.
Note that this may not be quite what is expected,
for instance, if there has been a back mutation to the ancestral allele at some point in the tree,
or if there have been mutations to distinct alleles on different parts of the tree
such that the ancestral allele is no longer present.
However, since these situations depend on multiple mutations occurring at a single site,
they are relatively rare in practice.

@@

\begin{figure}
    \centering
    \includegraphics{figures/branch_site_diagram}
    \caption{
    Computing the $f_4(A,B;C,D)$ statistic on a tree.
    \textbf{(a)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
    where $p_X$ is the proportion of the samples inheriting that mutation (colored stars) in set $X$
    (shown in labels at the bottom).
    Note that each of the mutations shown occurs at a distinct site, so there is no backmutation.
    \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
    to any mutation falling in it (colored lines), and then the branch $f_4$ statistic
    is the sum over these weights, multiplied by the length of the branch
    (and the span of the tree, not depicted).
        \label{fig:branch_site_diagram}
    }
\end{figure}

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome. We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
The \emph{polarised} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarised)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
However, the correspondence between the polarised statistics is less tight
than for the unpolarised ones.

@@

This relationship is true more generally.
In fact, for any region of the genome between $a$ and $b$,
with $\treeseq_{[a,b)}$ the trees spanning that interval,
\begin{align}
    \branch(f, \iw)_{[a,b)}
    =
    \frac{1}{\mu}
    \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
\end{align}
where the expected value leaves the genealogies fixed,
averaging over infinite-sites mutations
with rate $\mu$ per unit time and per unit of sequence length.
In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
which is itself a summary of local tree shapes.
Because we have normalised these statistics by the length of the genome under consideration,
both sides of this equation are in units of \emph{time}:
Branch statistics give mean weighted branch lengths;
Site statistics give mean densities of mutations per unit of sequence length,
which divided by $\mu$ is converted to time.
Let's unpack the assumptions here: what exactly is the mutational model?
First, we are taking the mutation rate to be $\mu$, i.e.,
the expected number of mutations that occur on a region of the genome of length $\ell$
over $t$ units of time is equal to $\mu \ell t$.
Second, we are assuming that the probability of per-site mutation is low enough
that no two mutations occur at the same site
--- the fact that they do, occasionally, means that this is an approximation.
Third, we are assuming that mutation rates are constant through time and across the genome.
Of course, the statement remains true if we can measure distance along the genome and time
in a way that mutation rates are constant, but how these vary is generally unknown.

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{harris2019database,kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}.
Specifically, we calculated diveristy using the site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
Is this an indication that Relate is doing a good job at inferring trees,
or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we describe and analyse the algorithm used
to compute Branch statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem shared across site, branch and node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted and removed.
By carefully ordering these insertions and removals,
we ensure that this state can be correctly maintained
using only small adjustments for each tree.
Computing the statistics is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropriately for the
particular statistics mode and windowing options.

@@

\begin{taocpalg}{B}{General branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$
compute the span-normalised statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.
}

\algstep{B1.}{Initialisation.}{
    For $0 \leq u < |\Et|$ set $\beta_u \leftarrow x_u
    \leftarrow F_u \leftarrow 0$ and $\pi_u \leftarrow -1$.
    Then, for $0 \leq j < n$ set $u \leftarrow S_j$,
    $x_u \leftarrow w_j$ and $F_u \leftarrow f(x_j)$. Finally,
    set $s \leftarrow \sigma \leftarrow j \leftarrow k \leftarrow t_l \leftarrow 0$.
}

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
(The initial state is the ``empty forest'', where no nodes are connected to any others.)
Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialisation in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous and tree are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ become disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each node $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

Once all of the edges with right coordinate equal to $t_l$ have been removed in
steps \algref{B3--B5}, we then insert edges that start in the current tree
(i.e., with left coordinate $t_l$) in steps \algref{B6--B8}. We update the
state to account for the new branch $u \mapsto v$ by setting the parent of
$u$ to $v$, computing the branch length of $u$ and adding the new contribution
of $u$ to $s$ in \algref{B7}. Then, as adding a new edge can connect the
subtree rooted at $u$ to the larger tree at $v$, we propagate the gain of $x_u$
up the tree in step \algref{B8} (note the symmetry with step \algref{B5}).
Finally, once we have removed and inserted all of the relevant edges, we are
ready to add the contribution of the current tree to the overall statistic,
$\sigma$ and move on to the next tree in step \algref{B9}. To do this, we first
compute the right-hand coordinate of the tree $t_r$ (a process slightly
complicated by the possible presence of ``gaps'' in the tree sequence, spanned
by no edges). Then, we add the running sum $s$ weighted by the span of the
current tree $t_r - t_l$ to $\sigma$, and return to \algref{B2} to process the
next tree. If we have reached the end of the sequence, we then divide $\sigma$
by the sequence length to normalise and exit.

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\begin{equation}\label{eqn:alg_complexity}
    O\left(n + \rho \log(n)^2 \right).
\end{equation}
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

@@

% duality: site stats ~ branch stat, but conversely branch more accurate version of site
%  maybe analogous to imputation
We hope a more broadly interesting and useful aspect of this paper
is a more in-depth look at the correspondence between statistics calculated with genome sequence
and properties of the underlying genealogies, i.e.,
the duality between Site and Branch statistics under the infinite-sites model.
This general correspondence is well-known,
but the common framework implemented here makes it easy to translate between the two.
One way to use this translation is to answer:
what aspect of tree shape is a particular population genetics statistic summarizing?
This can help when interpreting results, especially if reality may not fit an idealised model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to estimate the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
may use the signal from nearby patterns of variation
to interpret relationships at a given site,
thus transforming the simple binary split induced by a SNP
into a much richer source of information at every variant site.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.

@@

Multilocus statistics that measure linkage disequilibrium or haplotype lengths
do not fall in this category.
Extensions to statistics involving the pairwise joint distribution of genotypes across sites
\citep[as used by][]{hudson2001twolocus},
such as linkage disequlibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.

We also require that site statistics be computable from the genotype matrix
(i.e., without knowledge of the trees),
and, if the statistic is polarised, knowledge of the ancestral allele.
We often would also like our statistics to be unaffected by the addition of monomorphic sites,
which are usually omitted from genetic variation data.
To ensure this we have required that the summary function returns zero when give both zero weight
and the total weight: $f(0) = f(\tiw) = 0$.
This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
between any of the samples that have nonzero weight,
so that in particular,
\emph{simplification} of the tree sequence will not change the value of the statistic.
However, in practice this requirement can be relaxed.

@@

Let $h$ be a trait, $Z$ be a matrix of covariates, and $g$ be a vector denoting inheritance
(so, with $n$ samples, $h$ and $g$ are both $n$-vectors and $Z$ is an $n \times k$ matrix).
We would like to find the coefficient of $g$ in the linear regression of $h$ against $g$ and $Z$,
without doing full multivariate regression for every new $g$,
using the fact that $Z$ is always the same.
Suppose that $Z^T Z = I$ and that the vector of all ones is in the span of the columns of $Z$,
although in the implementation we post-process $Z$ to make this the case.
Then, let $a$ be the number and $b$ be the $k$-vector satisfying
\begin{align*}
    a, b = \text{argmin}\left\{ \sum_i \left( w_i - a g_i - \sum_j Z_{ij} b_j \right)^2 \right\}
\end{align*}
Writing this in block matrix notation, $a$ and $b$ minimise
\begin{align*}
    \left\|
        \left[ \begin{array}{@{}c@{}} h \end{array}\right]
            -
        \left[ \begin{array}{@{}c|c@{}} g & Z \end{array} \right]
            \left[ \begin{array}{@{}c@{}} a \\ \hline b \end{array} \right]
    \right\|^2 .
\end{align*}

@@

To compute this in the framework above,
first add a column of 1s to the covariates $Z$,
then decorrelate the resulting matrix, so that now $Z^T Z = I$.
Then, put this normalised version of $Z$
into the first $k$ columns of the weight matrix (so that $w_j(u) = Z_{uj}$),
set the $(k+1)^\text{st}$ column to the trait (so that $w_{k+1}(u) = w_u$),
and the final column to all $1$s (so $w_{k+2}(u) = 1$).
Also let $Z^T h = v$ be precomputed.
Then the sum of the traits of samples with the focal genotype is $h_g = x_{k+1}$,
and the allele count is $m = x_{k+2}$,
so that
\begin{align*}
    a
    &=
    \frac{
        h_g - \sum_{j=1}^k x_j v_j
    }{
        m - \sum_{j=1}^k x_j^2 } .
\end{align*}

>>>>

\paragraph{Polarization}
The node statistic defined above is appropriate for statistics such as nucleotide diversity, which is a function
of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
at a site, and we may define a \emph{polarized} node statistic without the second term from our previous definition:

\begin{align}
    \text{(polarized)} \qquad
    \nodep(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)) .
\end{align}

@@

In the definition above we sum over all alleles at each site.
However, sometimes it is useful to distinguish the \emph{ancestral} allele
(i.e., the allele at the root of the tree) from the remaining derived alleles.
This allows statistics in principle to differentiate ancestral from derived alleles,
information which is available in practice (albeit noisily),
and one way to make use of this information is to sum over only derived alleles.
Analogously to the above,
we say a site statistic is \textbf{polarized} if we do this,
defining
\begin{align} \label{eqn:site_polarised}
    \text{(polarized)} \qquad
    \sitep(f, \iw)_j
    &=
    \sum_{a \in D_j} f(\aw_j(a)) ,
\end{align}
where $D_j$ denotes the set of all alleles at site $j$ except the allele at the root of the tree.
Note that this may not be quite what is expected,
for instance, if there has been a back mutation to the ancestral allele at some point in the tree,
or if there have been mutations to distinct alleles on different parts of the tree
such that the ancestral allele is no longer present.
However, since these situations depend on multiple mutations occurring at a single site,
they are relatively rare in practice.

@@

\begin{figure}
    \centering
    \includegraphics{figures/branch_site_diagram}
    \caption{
    Computing the $f_4(A,B;C,D)$ statistic on a tree.
    \textbf{(a)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
    where $p_X$ is the proportion of the samples inheriting that mutation (colored stars) in set $X$
    (shown in labels at the bottom).
    Note that each of the mutations shown occurs at a distinct site, so there is no back mutation.
    \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
    to any mutation falling in it (colored lines), and then the branch $f_4$ statistic
    is the sum over these weights, multiplied by the length of the branch
    (and the span of the tree, not depicted).
        \label{fig:branch_site_diagram}
    }
\end{figure}

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome. We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
The \emph{polarized} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarized)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
However, the correspondence between the polarized statistics is less tight
than for the unpolished ones.

@@

This relationship is true more generally.
In fact, for any region of the genome between $a$ and $b$,
with $\treeseq_{[a,b)}$ the trees spanning that interval,
\begin{align}
    \branch(f, \iw)_{[a,b)}
    =
    \frac{1}{\mu}
    \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
\end{align}
where the expected value leaves the genealogies fixed,
averaging over infinite-sites mutations
with rate $\mu$ per unit time and per unit of sequence length.
In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
which is itself a summary of local tree shapes.
Because we have normalized these statistics by the length of the genome under consideration,
both sides of this equation are in units of \emph{time}:
Branch statistics give mean weighted branch lengths;
Site statistics give mean densities of mutations per unit of sequence length,
which divided by $\mu$ is converted to time.
Let's unpack the assumptions here: what exactly is the mutational model?
First, we are taking the mutation rate to be $\mu$, i.e.,
the expected number of mutations that occur on a region of the genome of length $\ell$
over $t$ units of time is equal to $\mu \ell t$.
Second, we are assuming that the probability of per-site mutation is low enough
that no two mutations occur at the same site
--- the fact that they do, occasionally, means that this is an approximation.
Third, we are assuming that mutation rates are constant through time and across the genome.
Of course, the statement remains true if we can measure distance along the genome and time
in a way that mutation rates are constant, but how these vary is generally unknown.

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{harris2019database,kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}.
Specifically, we calculated diversity using the site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
Is this an indication that Relate is doing a good job at inferring trees,
or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we describe and analyze the algorithm used
to compute Branch statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem shared across site, branch and node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted and removed.
By carefully ordering these insertions and removals,
we ensure that this state can be correctly maintained
using only small adjustments for each tree.
Computing the statistics is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropriately for the
particular statistics mode and windowing options.

@@

\begin{taocpalg}{B}{General branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$
compute the span-normalized statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.
}

\algstep{B1.}{Initialization.}{
    For $0 \leq u < |\Et|$ set $\beta_u \leftarrow x_u
    \leftarrow F_u \leftarrow 0$ and $\pi_u \leftarrow -1$.
    Then, for $0 \leq j < n$ set $u \leftarrow S_j$,
    $x_u \leftarrow w_j$ and $F_u \leftarrow f(x_j)$. Finally,
    set $s \leftarrow \sigma \leftarrow j \leftarrow k \leftarrow t_l \leftarrow 0$.
}

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
(The initial state is the ``empty forest'', where no nodes are connected to any others.)
Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialization in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous and tree are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ become disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each node $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

Once all of the edges with right coordinate equal to $t_l$ have been removed in
steps \algref{B3--B5}, we then insert edges that start in the current tree
(i.e., with left coordinate $t_l$) in steps \algref{B6--B8}. We update the
state to account for the new branch $u \mapsto v$ by setting the parent of
$u$ to $v$, computing the branch length of $u$ and adding the new contribution
of $u$ to $s$ in \algref{B7}. Then, as adding a new edge can connect the
subtree rooted at $u$ to the larger tree at $v$, we propagate the gain of $x_u$
up the tree in step \algref{B8} (note the symmetry with step \algref{B5}).
Finally, once we have removed and inserted all of the relevant edges, we are
ready to add the contribution of the current tree to the overall statistic,
$\sigma$ and move on to the next tree in step \algref{B9}. To do this, we first
compute the right-hand coordinate of the tree $t_r$ (a process slightly
complicated by the possible presence of ``gaps'' in the tree sequence, spanned
by no edges). Then, we add the running sum $s$ weighted by the span of the
current tree $t_r - t_l$ to $\sigma$, and return to \algref{B2} to process the
next tree. If we have reached the end of the sequence, we then divide $\sigma$
by the sequence length to normalize and exit.

To analyze Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\begin{equation}\label{eqn:alg_complexity}
    O\left(n + \rho \log(n)^2 \right).
\end{equation}
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

@@

% duality: site stats ~ branch stat, but conversely branch more accurate version of site
%  maybe analogous to imputation
We hope a more broadly interesting and useful aspect of this paper
is a more in-depth look at the correspondence between statistics calculated with genome sequence
and properties of the underlying genealogies, i.e.,
the duality between Site and Branch statistics under the infinite-sites model.
This general correspondence is well-known,
but the common framework implemented here makes it easy to translate between the two.
One way to use this translation is to answer:
what aspect of tree shape is a particular population genetics statistic summarizing?
This can help when interpreting results, especially if reality may not fit an idealized model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to estimate the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
may use the signal from nearby patterns of variation
to interpret relationships at a given site,
thus transforming the simple binary split induced by a SNP
into a much richer source of information at every variant site.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.

@@

Multilocus statistics that measure linkage disequilibrium or haplotype lengths
do not fall in this category.
Extensions to statistics involving the pairwise joint distribution of genotypes across sites
\citep[as used by][]{hudson2001twolocus},
such as linkage disequilibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.

We also require that site statistics be computable from the genotype matrix
(i.e., without knowledge of the trees),
and, if the statistic is polarized, knowledge of the ancestral allele.
We often would also like our statistics to be unaffected by the addition of monomorphic sites,
which are usually omitted from genetic variation data.
To ensure this we have required that the summary function returns zero when give both zero weight
and the total weight: $f(0) = f(\tiw) = 0$.
This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
between any of the samples that have nonzero weight,
so that in particular,
\emph{simplification} of the tree sequence will not change the value of the statistic.
However, in practice this requirement can be relaxed.

@@

Let $h$ be a trait, $Z$ be a matrix of covariates, and $g$ be a vector denoting inheritance
(so, with $n$ samples, $h$ and $g$ are both $n$-vectors and $Z$ is an $n \times k$ matrix).
We would like to find the coefficient of $g$ in the linear regression of $h$ against $g$ and $Z$,
without doing full multivariate regression for every new $g$,
using the fact that $Z$ is always the same.
Suppose that $Z^T Z = I$ and that the vector of all ones is in the span of the columns of $Z$,
although in the implementation we post-process $Z$ to make this the case.
Then, let $a$ be the number and $b$ be the $k$-vector satisfying
\begin{align*}
    a, b = \text{argmin}\left\{ \sum_i \left( w_i - a g_i - \sum_j Z_{ij} b_j \right)^2 \right\}
\end{align*}
Writing this in block matrix notation, $a$ and $b$ minimize
\begin{align*}
    \left\|
        \left[ \begin{array}{@{}c@{}} h \end{array}\right]
            -
        \left[ \begin{array}{@{}c|c@{}} g & Z \end{array} \right]
            \left[ \begin{array}{@{}c@{}} a \\ \hline b \end{array} \right]
    \right\|^2 .
\end{align*}

@@

To compute this in the framework above,
first add a column of 1s to the covariates $Z$,
then decorrelate the resulting matrix, so that now $Z^T Z = I$.
Then, put this normalized version of $Z$
into the first $k$ columns of the weight matrix (so that $w_j(u) = Z_{uj}$),
set the $(k+1)^\text{st}$ column to the trait (so that $w_{k+1}(u) = w_u$),
and the final column to all $1$s (so $w_{k+2}(u) = 1$).
Also let $Z^T h = v$ be precomputed.
Then the sum of the traits of samples with the focal genotype is $h_g = x_{k+1}$,
and the allele count is $m = x_{k+2}$,
so that
\begin{align*}
    a
    &=
    \frac{
        h_g - \sum_{j=1}^k x_j v_j
    }{
        m - \sum_{j=1}^k x_j^2 } .
\end{align*}
